Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-30 paper.tex
Updates from Overleaf

The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
on the number of events involving trapped ancestral material in Hudson's algorithm.
For $n$ samples from
a diploid population of effective size $N_e$,
and a genome of genetic length $r$,
with $\rho = 4 N_e r$ the scaled recombination rate,
and $R^T$ the number of recombination recombination events in trapped material,
this upper bound is
\begin{equation} \label{eqn-hsw-bound}
    \mathbb{E}[R^T]
    \le
    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
\end{equation}
Note that the sum on the right is well-approximated by $\log(n)$.
Although \citet{hein2004gene} refer to this bound as ``crude'',
it turns out to predict the scaling of runtime quite well for large genomes.
Figure~\ref{fig-expected-perf} shows the observed runtime
(from the same simulations as Figure~\ref{fig-ancestry-perf})
plotted against equation~\eqref{eqn-hsw-bound},
showing that for values of $\rho$ above $4 \times 10^4$ or so
(i.e., chromosome-length simulations in populations above $10^4$),
the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
This implies the runtime for a large sample
can be predicted from the runtime for a small sample.
For smaller values of $\rho$, however, the prediction
severely underestimates runtime.

>>>>

The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
on the number of events involving trapped ancestral material in Hudson's algorithm.
For $n$ samples from
a diploid population of effective size $N_e$,
and a genome of genetic length $r$,
with $\rho = 4 N_e r$ the scaled recombination rate,
and $R^T$ the number of recombination events in trapped material,
this upper bound is
\begin{equation} \label{eqn-hsw-bound}
    \mathbb{E}[R^T]
    \le
    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
\end{equation}
Note that the sum on the right is well-approximated by $\log(n)$.
Although \citet{hein2004gene} refer to this bound as ``crude'',
it turns out to predict the scaling of runtime quite well for large genomes.
Figure~\ref{fig-expected-perf} shows the observed runtime
(from the same simulations as Figure~\ref{fig-ancestry-perf})
plotted against equation~\eqref{eqn-hsw-bound},
showing that for values of $\rho$ above $4 \times 10^4$ or so
(i.e., chromosome-length simulations in populations above $10^4$),
the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
This implies the runtime for a large sample
can be predicted from the runtime for a small sample.
For smaller values of $\rho$, however, the prediction
severely underestimates runtime.
